### 2024-10-11 manuscript.tex
grammar [ci skip]
This build is based on
https://github.com/pivlab/annual_review_of_biomedical_data_science/commit/6f7af112710ae5f8179f2e9d88b13533a8cab962.

This commit was created by the following CI build and job:
https://github.com/pivlab/annual_review_of_biomedical_data_science/commit/6f7af112710ae5f8179f2e9d88b13533a8cab962/checks
https://github.com/pivlab/annual_review_of_biomedical_data_science/actions/runs/11283899150

This manuscript
(\href{https://pivlab.github.io/annual_review_of_biomedical_data_science/v/4e9927114201ae1607c8d2e8ce90c7a85fe84773/}{permalink})
was automatically generated
from \href{https://github.com/pivlab/annual_review_of_biomedical_data_science/tree/4e9927114201ae1607c8d2e8ce90c7a85fe84773}{pivlab/annual\_review\_of\_biomedical\_data\_science@4e99271}
on October 10, 2024.

@@

Understanding the genetics of complex traits remains one of the key challenges in modern genomics.
Genome-wide association studies (GWAS) have significantly advanced our knowledge by identifying associations between genetic variants and traits, shedding light on the genetic architecture underlying a variety of diseases and phenotypic conditions.
Since the first GWAS on age-related macular degeneration in 2005, the GWAS catalog has grown to include over 5,000 human traits, encompassing risk loci for conditions such as type 2 diabetes, schizophrenia, and rheumatoid arthritis (\protect\hyperlink{ref-HicyXvA4}{3}).
However, most GWAS-identified variants reside in non-coding regions, underscoring the importance of gene regulation in generating phenotypic diversity (\protect\hyperlink{ref-BqjxxhyF}{4}, \protect\hyperlink{ref-9Pr9idng}{5}).
This gap presents a major challenge in translating GWAS findings into biological insights.

To address this limitation, numerous methods have been developed to link GWAS variants to functional outcomes.
Proximity-based methods, such as MAGMA, have been widely used to link GWAS variants to nearby genes based on physical distance (\protect\hyperlink{ref-19XiXgYmd}{6}).
However, these methods may not consider long-range acting SNPs, such as the variants in the \emph{FTO} locus that are associated with obesity, which have been shown to influence the \emph{IRX3} gene through long-range interactions (\protect\hyperlink{ref-167QL5tMV}{7}).
While useful, proximity-based approaches may miss causal genes if they are not the closest ones.
Transcriptome-wide association studies (TWAS) have proven effective by linking genetic variants to gene expression, thereby enhancing our ability to identify putatively causal genes for various traits (\protect\hyperlink{ref-l6ogswV3}{9}).
TWAS employs expression quantitative trait loci (eQTL) data to prioritize genes that are influenced by GWAS variants, offering a more mechanistic link between genetic variation and phenotypic changes.
Nevertheless, TWAS and similar methods typically focus on individual genes or variants, overlooking the complex interactions among genes that are increasingly recognized as vital to understanding complex traits (\protect\hyperlink{ref-vpIDZCSa}{10}).

@@

\hypertarget{single-variant-single-gene-approaches}{%
\subsection{Single variant / single gene approaches}\label{single-variant-single-gene-approaches}}

\begin{figure}
\hypertarget{fig:fig1}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/fig1.svg}
\caption{\textbf{Major approaches based on single variant/single gene strategies to understand the genetic basis of complex traits}.
GWAS identifies associations between DNA variants and traits.
All the other approaches test how different molecular mechanisms might mediate variant-trait associations from GWAS.
TWAS uses variants linked to changes in gene expression (i.e., expression quantitative trait loci or eQTLs).
PWAS links protein functionality to traits via protein QTLs (pQTLs).
HAWAS maps histone modifications to traits using hQTLs, while MWAS explores DNA methylation patterns in relation to traits through mQTLs.}\label{fig:fig1}
}
\end{figure}

@@

In a simple trait, a single gene can be responsible for the disease (monogenic), for example in the case of Sickle Cell Anemia and Huntington's disease.
However, most traits are not so simplistic and are the result of many mutations across the genome (polygenic).
Even monogenic traits can sometimes be influenced by the polygenic background (\protect\hyperlink{ref-VqGYQMXO}{14}).
Genome-wide association studies (GWAS) examine the relationship between specific genetic variants and phenotypes by comparing allele frequencies in individuals of similar ancestry with distinct phenotypic traits (\textbf{Figure \ref{fig:fig1}}).
Both causal and non-causal variants are found to be significant and due to the phenomena of linkage disequilibrium, the results of GWAS are often grouped into risk loci.
These loci represent clusters of variants that demonstrate significant associations, although not every variant may be causal (\protect\hyperlink{ref-1873zCaaX}{15}).

Since the first GWAS on age-related macular degeneration (AMD) in 2005, the GWAS catalog rapidly expanded, containing SNP-trait associations across \textgreater5,000 human traits (\protect\hyperlink{ref-VRCjjZ9L}{1}).
These studies have successfully identified risk loci for a variety of traits such as type 2 diabetes, auto-immune disease, schizophrenia, major depressive disorder, and many more (\protect\hyperlink{ref-17B4AigZT}{2}, \protect\hyperlink{ref-yqBcaXHD}{16}).
As GWAS participants increase, more loci are able to be identified.
For example, a study investigating insomnia with a sample size of \textgreater1M identified 202 risk loci (\protect\hyperlink{ref-sBTrhvlJ}{17}), compared to earlier studies with a sample size of \textasciitilde110,00, which were only able to identify 3 risk loci (\protect\hyperlink{ref-crCAymA1}{18}).
Beyond the identification of risk loci, GWAS have also led to the discovery of novel biological mechanisms in Crohn's disease (\protect\hyperlink{ref-s6z5ZuY3}{19}, \protect\hyperlink{ref-gxwza7fO}{20}), rheumatoid arthritis (\protect\hyperlink{ref-HicyXvA4}{3}), and more.

Although GWAS have identified numerous genetic variants associated with complex traits, translating these findings into biological insights remains challenging (\protect\hyperlink{ref-17B4AigZT}{2}).
For example, genes can have epistatic interactions where a secondary loci affects a primary loci.
The peripheral effects of the secondary loci are often too small to be picked up through GWAS alone (\protect\hyperlink{ref-IrGVpNji}{21}).
Furthermore, any significant GWAS variants are located in non-coding regions, making it difficult to identify the specific genes that drive trait associations (\protect\hyperlink{ref-BqjxxhyF}{4}).
A simple strategy to link GWAS-associated variants with a gene is by physical proximity, which typically selects the closest gene to the most significant SNP at each locus.
Approaches like MAGMA (\protect\hyperlink{ref-19XiXgYmd}{6}) are based on proximity to compute a gene-trait association.
However, SNPs can have distal effects on the expression of the target gene (\protect\hyperlink{ref-9Pr9idng}{5}, \protect\hyperlink{ref-167QL5tMV}{7}, \protect\hyperlink{ref-OBxYzUMh}{22}, \protect\hyperlink{ref-WfdUHVDA}{23}).

Recent approaches combine multi-omics data from various cell types and tissues with GWAS to identify potential mechanisms of SNPs and the associated gene through molecular quantitative trait loci (molQTLs) (\protect\hyperlink{ref-QIEm5ZWK}{24}) (Figure \ref{fig:fig1}).
For example, an expression quantitative trait loci (eQTL) is a genetic region associated with the expression levels of a nearby or distant gene (\protect\hyperlink{ref-1FHKg3JK5}{25}).
eQTL studies estimate that the majority of heritability is explained by the combination of multiple weak \emph{trans}-eQTLs (\protect\hyperlink{ref-sGgkLsyy}{26}).

@@

Transcriptome-wide association studies (TWAS) integrate GWAS with gene expression data (\textbf{Figure \ref{fig:fig1}}) from eQTL analysis, to prioritize genes whose expression across different tissues is influenced by GWAS variants (\protect\hyperlink{ref-ReOPt75u}{8}, \protect\hyperlink{ref-l6ogswV3}{9}).
By leveraging predicted gene expression levels, TWAS provides a mechanistic link between genetic variants and traits, allowing researchers to move beyond associations with individual SNPs to identify putatively causal genes (\protect\hyperlink{ref-l6ogswV3}{9}).
Since TWAS models the genetic regulation of gene expression, the approach allows researchers to impute expression levels in GWAS cohorts where expression data may not be available.
A key advantage of TWAS over GWAS lies in its ability to increase interpretability by providing a gene-trait association: TWAS connects trait-associated SNPs (which are mostly non-coding) to genes, which are biologically functional units.

@@

Epigenome-wide association studies (EWAS) encompass methodologies such methylome-wide association studies (MWAS) (\textbf{Figure \ref{fig:fig1}}) and histone acetylome-wide association studies (HAWAS) (\textbf{Figure \ref{fig:fig1}}), based on haQTLs and mQLTs respectability, which focus on specific types of epigenetic modifications to reveal their roles in gene regulation and disease etiology (\protect\hyperlink{ref-TPqvAJda}{35}).
MWAS targets DNA methylation patterns, identifying loci where methylation changes are linked to particular traits or diseases.
For instance, a study by Shen et al.~demonstrated a causal relationship between DNA methylation and depression, indicating that epigenetic modifications may mediate genetic risk for psychiatric disorders (\protect\hyperlink{ref-1DNxNkLsT}{36}).
HAWAS focuses on histone acetylation modifications, which are crucial for regulating chromatin structure and gene expression.
Del Rosario et al.conducted a HAWAS identifying over 2,000 differentially acetylated loci in immune cells from Mycobacterium tuberculosis-infected individuals, linking these changes to gene expression and potassium channel genes like \emph{KCNJ15} in modulating apoptosis and Mtb clearance.
haQTL analysis further revealed variants associated with immune phenotypes, complementing GWAS findings and enhancing understanding of disease mechanisms (\protect\hyperlink{ref-1BYHeZlgB}{37}).
The tissue-specific nature of epigenetic modifications, alongside the capacity to capture cellular plasticity and environmental influences, enhances our insight into the effects of genetic variants across distinct tissues, temporal cellular states, and gene-environment interactions (\protect\hyperlink{ref-FeaSnwhX}{38}, \protect\hyperlink{ref-KfWCLspO}{39}).

@@

Li et al.~(2024) identified a significantly connected subgraph formed by cancer-affected coding genes and ncRNAs by focusing on connectivity as a key topological feature.
This approach highlighted the essential role of ncRNAs in linking fragmented cancer-affected genes, consistent with the omnigenic framework's premise that peripheral genes, including non-coding elements, contribute to complex traits such as cancer.
Importantly, the inclusion of ncRNAs enhanced the identification of cancer-related pathways, indicating that a comprehensive network model encompassing ncRNAs is more effective in characterizing disease relationships than models concentrating solely on coding genes (\protect\hyperlink{ref-13VkGrWQR}{50}).
Empirical support for the omnigenic theory extends beyond human diseases.
Sundell et al.~demonstrated this concept in European black poplars, showing that both core and peripheral genes play a crucial role in predicting phenotypes.
While core genes are indeed important, the information they provide must be complemented by other genes to ensure accurate phenotype predictions.
Furthermore, peripheral genes were found to carry significant biological information contributing to robust predictions (\protect\hyperlink{ref-EgquVUSg}{51}).

@@

In essence, the omnigenic model proposes a specific mechanism: that the genetic architecture of complex traits is intrinsically interconnected, involving a multitude of genes that contribute to phenotypic variation both directly (core genes) and indirectly (peripheral genes).
This framework facilitates an understanding of the highly polygenic and often subtle genetic effects on complex traits, shifting the focus from individual genes to gene networks.
While this proves successful in some fields the model exhibits certain limitations and has faced several criticisms.

@@

Furthermore, the omnigenic model has been criticized for their binary classification of genes into core and peripheral, which might oversimplify biological systems and potentially underestimate their true complexity (\protect\hyperlink{ref-82fs3SZo}{52}).
It also fails to account for gene-environment interactions, which play a crucial role in shaping complex traits (\protect\hyperlink{ref-dg9nKuy0}{55}).
Additionally, while the model provides a conceptual framework, it remains unclear how to translate it into a practical statistical model (\protect\hyperlink{ref-1GlQtJ9Yz}{56}).

@@

\begin{figure}
\hypertarget{fig:fig3}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/fig3.svg}
\caption{\textbf{Computational methods for machine learning derived gene module detection.}.
\emph{(a)} Real biological network, where nodes represent genes and edges denote the relationships between them.
It captures all possible patterns that can link genes.
Some of these relationships are gene co-expression patterns, which transcriptomics can capture, such as (1), (3) and (4).
Other patterns, such as protein-protein interactions (2), are not captured by transcriptomics and are thus not reflected in the co-expression analysis.
\emph{(b)} Different approaches for capturing machine learning-derived gene modules.
In this example, transcriptome data is used as input, but other data types such as methylomics, proteomics, and metabolomics can also be applied.
The first row illustrates the PLIER method, which uses non-negative matrix factorization and prior knowledge to align gene modules with known biological pathways.
This approach generates interpretable gene modules but only captures linear relationships.
The second row shows a vanilla VAE, which can capture both linear and nonlinear patterns but does not integrate prior knowledge, resulting in less interpretable gene modules.
The third row highlights interpretable VAEs, which incorporate prior knowledge into its decoder, achieving a balance between capturing complex relationships and maintaining biological interpretability.}\label{fig:fig3}
}
\end{figure}

@@

GenomicSuperSignature employs PCA to identify gene modules applying prior knowledge for biological annotation after module discovery.
This flexibility allows it to adapt to novel datasets and provides computational efficiency, even with large datasets.
However, the lack of prior knowledge integration during training can result in modules that are less biologically interpretable compared to those from PLIER-based methods (\protect\hyperlink{ref-X4fhSCkz}{65}).

@@

DL models are also highly effective in unsupervised settings such as inferring gene modules.
Specifically, DL methods like autoencoders (\protect\hyperlink{ref-NLVTJ9Lj}{73}), which utilize neural networks to learn compressed representations, can capture both linear and nonlinear patterns between genes.
This makes autoencoders particularly suited for uncovering gene modules and efficiently handling high-dimensional datasets.
For instance, variational autoencoders (VAEs) (\textbf{Figure \ref{fig:fig3}\emph{b}}) enhance this process by encoding data into a continuous latent space, enabling dimensionality reduction and the generation of new data.
However, this increased power to capture more general patterns has a major tradeoff: interpretability.
To overcome this, interpretable VAE models like the Pathway Module VAE (pmVAE) (\protect\hyperlink{ref-nGVsOd2S}{74}), expiMap (\protect\hyperlink{ref-JkqcQgM7}{75}), and VEGA (\protect\hyperlink{ref-170T6ip47}{12}) integrate biological pathway information into the decoder, which improves interpretability, thus making these approaches function as linear dimensionality reduction methods.
(\textbf{Figure \ref{fig:fig3}\emph{b}}).

@@

\begin{figure}
\hypertarget{fig:fig4}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/fig4.svg}
\caption{\textbf{An integrative, gene module-based approach for genetic studies.}
Each panel shows a component of the PhenoPLIER framework (\protect\hyperlink{ref-NM3rHx1i}{13}).
\emph{(a)} First, latent variables (LVs) or gene modules are learned from transcriptome data using the PLIER matrix factorization method.
PLIER generates matrix \(\mathbf{Z}\), which has gene weights for each module, and matrix \(\mathbf{B}\), which has the samples in the latent space.
\emph{(b)} Schematic of gene-trait associations from TWAS and gene-drug scores from LINCS L1000 being projected into the latent space for a joint analysis.
\emph{(c)} Schematic of a gene module-based drug reporposing framework, where the projection of TWAS and LINCS L1000 data is used to compute a drug-disease score.
\emph{(d)} Schematic of a regression model that tests whether genes that belong to a module (using a column of \(\mathbf{Z}\)) tend to be more strongly associated with a trait (using \(p\)-values from TWAS).
\emph{(e)} (top) Example of a gene module identified as LV246 analyzed in (\protect\hyperlink{ref-NM3rHx1i}{13}).
\emph{DGAT2} and \emph{ACACA}, found in a CRISPR screen to be linked to lipid metabolism and potential core genes, were among the top 15 genes in this module.
Using metadata from the top samples for LV246 from \(\mathbf{B}\), the module was found to be expressed mainly in adipose tissue, and liver and astrocyes were part of the top 10 cell types and tissues.
Using the regression model depicted in d), this module was significantly associated with high cholesterol, tryglycerides, LDL cholesterol, cholesterol lowering medication, Alzheimer's disease and dementia in the discovery cohort (UK Biobank), and high cholesterol and memory loss in the replication cohort (eMERGE).
(bottom) The \(p\)-values and colocalization probability from gene-trait associations from TWAS for a subset of the top 1\% genes in LV246 (\(y\)-axis) and traits significantly associated with LV246 (\(x\)-axis).
It can be seen that \emph{DGAT2} and \emph{ACACA} (in boldface), although strongly linked to lipid metabolism, are weakly associated with these lipid-relevant traits.}\label{fig:fig4}
}
\end{figure}

All the previously described approaches provide valuable and orthogonal information to better understand the molecular basis of human complex traits, including different molecular mechanisms through variant-trait, variant-gene, gene-trait and gene-gene information.
All of them have limitations and offer a partial view of the complex molecular interplay underlying pathophysiological processes.
In this section, we describe a gene module-based approach aimed at identifying patterns shared across these data sources.
The underlying hypothesis of this approach is that biologically meaningful, disease-relevant gene-gene links will be revealed when different yet incomplete data modalities are integrated.
In particular, we show examples on how the integration of gene modules with gene-trait associations and gene-drug data highlight core genes, disease-relevant molecular processes and drug's mechanisms of action.

>>>>

This manuscript
(\href{https://pivlab.github.io/annual_review_of_biomedical_data_science/v/6f7af112710ae5f8179f2e9d88b13533a8cab962/}{permalink})
was automatically generated
from \href{https://github.com/pivlab/annual_review_of_biomedical_data_science/tree/6f7af112710ae5f8179f2e9d88b13533a8cab962}{pivlab/annual\_review\_of\_biomedical\_data\_science@6f7af11}
on October 11, 2024.

@@

Understanding the genetics of complex traits remains one of the key challenges in modern genomics.
GWAS have significantly advanced our knowledge by identifying associations between genetic variants and traits, shedding light on the genetic architecture underlying a variety of diseases and phenotypic conditions.
Since the first GWAS on age-related macular degeneration in 2005, the GWAS catalog has grown to include over 5,000 human traits, encompassing risk loci for conditions such as type 2 diabetes, schizophrenia, and rheumatoid arthritis (\protect\hyperlink{ref-HicyXvA4}{3}).
However, most GWAS-identified variants reside in non-coding regions, underscoring the importance of gene regulation in generating phenotypic diversity (\protect\hyperlink{ref-BqjxxhyF}{4}, \protect\hyperlink{ref-9Pr9idng}{5}).
This gap presents a major challenge in translating GWAS findings into biological insights.

To address this limitation, numerous methods have been developed to link GWAS variants to functional outcomes.
Proximity-based methods, such as MAGMA, have been widely used to link GWAS variants to nearby genes based on physical distance (\protect\hyperlink{ref-19XiXgYmd}{6}).
However, these methods may not consider long-range acting SNPs, such as the variants in the \emph{FTO} locus that are associated with obesity, which have been shown to influence the \emph{IRX3} gene through long-range interactions (\protect\hyperlink{ref-167QL5tMV}{7}).
While useful, proximity-based approaches may miss causal genes if they are not the closest ones.
TWAS have proven effective by linking genetic variants to gene expression, thereby enhancing our ability to identify putatively causal genes for various traits (\protect\hyperlink{ref-l6ogswV3}{9}).
TWAS employs expression quantitative trait loci (eQTL) data to prioritize genes that are influenced by GWAS variants, offering a more mechanistic link between genetic variation and phenotypic changes.
Nevertheless, TWAS and similar methods typically focus on individual genes or variants, overlooking the complex interactions among genes that are increasingly recognized as vital to understanding complex traits (\protect\hyperlink{ref-vpIDZCSa}{10}).

@@

\hypertarget{single-variant-and-single-gene-approaches}{%
\subsection{Single variant and single gene approaches}\label{single-variant-and-single-gene-approaches}}

\begin{figure}
\hypertarget{fig:fig1}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/fig1.svg}
\caption{\textbf{Major approaches based on single variant and single gene strategies to understand the genetic basis of complex traits}.
Genome-wide association studies (GWAS) identify associations between DNA variants and traits.
All the other approaches test how different molecular mechanisms might mediate variant-trait associations from GWAS.
Transcriptome-wide association studies (TWAS) use variants linked to changes in gene expression (i.e., expression quantitative trait loci or eQTLs).
Protein-wide association studies (PWAS) links protein functionality to traits via protein QTLs (pQTLs).
Histone acetylome-wide association studies (HAWAS) maps histone acetylation modifications to traits using histone acetylation QTLs (haQTLs), while methylation-wide association studies (MWAS) explores DNA methylation patterns in relation to traits through methylation QTLs (mQTLs).}\label{fig:fig1}
}
\end{figure}

@@

In a simple trait, a single gene can be responsible for the disease (monogenic), for example in the case of Sickle Cell Anemia and Huntington's disease.
However, most traits are not so simplistic and are the result of many mutations across the genome (polygenic).
Even monogenic traits can sometimes be influenced by the polygenic background (\protect\hyperlink{ref-VqGYQMXO}{14}).
GWAS examine the relationship between specific genetic variants and phenotypes by comparing allele frequencies in individuals of similar ancestry with distinct phenotypic traits (\textbf{Figure \ref{fig:fig1}}).
Both causal and non-causal variants are found to be significant and due to the phenomenon of linkage disequilibrium, the results of GWAS are often grouped into risk loci.
These loci represent clusters of variants that demonstrate significant associations, although not every variant may be causal (\protect\hyperlink{ref-1873zCaaX}{15}).

Since the first GWAS on age-related macular degeneration (AMD) in 2005, the GWAS catalog has rapidly expanded, containing SNP-trait associations across \textgreater5,000 human traits (\protect\hyperlink{ref-VRCjjZ9L}{1}).
These studies have successfully identified risk loci for a variety of traits such as type 2 diabetes, auto-immune disease, schizophrenia, major depressive disorder, and many more (\protect\hyperlink{ref-17B4AigZT}{2}, \protect\hyperlink{ref-yqBcaXHD}{16}).
As GWAS participants increase, more loci are able to be identified.
For example, a study investigating insomnia with a sample size of \textgreater1M identified 202 risk loci (\protect\hyperlink{ref-sBTrhvlJ}{17}), compared to earlier studies with a sample size of \textasciitilde110,00, which were only able to identify 3 risk loci (\protect\hyperlink{ref-crCAymA1}{18}).
Beyond the identification of risk loci, GWAS have also led to the discovery of novel biological mechanisms in Crohn's disease (\protect\hyperlink{ref-s6z5ZuY3}{19}, \protect\hyperlink{ref-gxwza7fO}{20}), rheumatoid arthritis (\protect\hyperlink{ref-HicyXvA4}{3}), and more.

Although GWAS have identified numerous genetic variants associated with complex traits, translating these findings into biological insights remains challenging (\protect\hyperlink{ref-17B4AigZT}{2}).
For example, genes can have epistatic interactions where a secondary loci affects a primary locus.
The peripheral effects of the secondary loci are often too small to be picked up through GWAS alone (\protect\hyperlink{ref-IrGVpNji}{21}).
Furthermore, any significant GWAS variants are located in non-coding regions, making it difficult to identify the specific genes that drive trait associations (\protect\hyperlink{ref-BqjxxhyF}{4}).
A simple strategy to link GWAS-associated variants with a gene is by physical proximity, which typically selects the closest gene to the most significant SNP at each locus.
Approaches like MAGMA (\protect\hyperlink{ref-19XiXgYmd}{6}) are based on proximity to compute a gene-trait association.
However, SNPs can have distal effects on the expression of the target gene (\protect\hyperlink{ref-9Pr9idng}{5}, \protect\hyperlink{ref-167QL5tMV}{7}, \protect\hyperlink{ref-OBxYzUMh}{22}, \protect\hyperlink{ref-WfdUHVDA}{23}).

Recent approaches combine multi-omics data from various cell types and tissues with GWAS to identify potential mechanisms of SNPs and the associated gene through molecular quantitative trait loci (molQTLs) (\protect\hyperlink{ref-QIEm5ZWK}{24}) (Figure \ref{fig:fig1}).
For example, an expression quantitative trait locus (eQTL) is a genetic region associated with the expression levels of a nearby or distant gene (\protect\hyperlink{ref-1FHKg3JK5}{25}).
eQTL studies estimate that the majority of heritability is explained by the combination of multiple weak \emph{trans}-eQTLs (\protect\hyperlink{ref-sGgkLsyy}{26}).

@@

Transcriptome-wide association studies (TWAS) integrate GWAS with gene expression data (\textbf{Figure \ref{fig:fig1}}) from eQTL analysis, to prioritize genes whose expression across different tissues is influenced by GWAS variants (\protect\hyperlink{ref-ReOPt75u}{8}, \protect\hyperlink{ref-l6ogswV3}{9}).
By leveraging predicted gene expression levels, TWAS provides a mechanistic link between genetic variants and traits, allowing researchers to move beyond associations with individual SNPs to identify putatively causal genes (\protect\hyperlink{ref-l6ogswV3}{9}).
Since TWAS models the genetic regulation of gene expression, this approach enables researchers to impute expression levels in GWAS cohorts where expression data may not be available.
A key advantage of TWAS over GWAS lies in its ability to increase interpretability by providing a gene-trait association: TWAS connects trait-associated SNPs (which are mostly non-coding) to genes, which are biologically functional units.

@@

Epigenome-wide association studies (EWAS) encompass methodologies such as methylome-wide association studies (MWAS) (\textbf{Figure \ref{fig:fig1}}) and histone acetylome-wide association studies (HAWAS) (\textbf{Figure \ref{fig:fig1}}), based on haQTLs and mQLTs respectability, which focus on specific types of epigenetic modifications to reveal their roles in gene regulation and disease etiology (\protect\hyperlink{ref-TPqvAJda}{35}).
MWAS targets DNA methylation patterns, identifying loci where methylation changes are linked to particular traits or diseases.
For instance, a study by Shen et al.~demonstrated a causal relationship between DNA methylation and depression, indicating that epigenetic modifications may mediate genetic risk for psychiatric disorders (\protect\hyperlink{ref-1DNxNkLsT}{36}).
HAWAS focuses on histone acetylation modifications, which are crucial for regulating chromatin structure and gene expression.
Del Rosario et al.conducted a HAWAS identifying over 2,000 differentially acetylated loci in immune cells from Mycobacterium tuberculosis-infected individuals, linking these changes to gene expression and potassium channel genes like \emph{KCNJ15} in modulating apoptosis and Mtb clearance.
haQTL analysis further revealed variants associated with immune phenotypes, complementing GWAS findings and enhancing understanding of disease mechanisms (\protect\hyperlink{ref-1BYHeZlgB}{37}).
The tissue-specific nature of epigenetic modifications, alongside the capacity to capture cellular plasticity and environmental influences, enhances our insight into the effects of genetic variants across distinct tissues, temporal cellular states, and gene-environment interactions (\protect\hyperlink{ref-FeaSnwhX}{38}, \protect\hyperlink{ref-KfWCLspO}{39}).

@@

Li et al.~(2024) identified a significantly connected subgraph formed by cancer-affected coding genes and ncRNAs by focusing on connectivity as a key topological feature.
This approach highlighted the essential role of ncRNAs in linking fragmented cancer-affected genes, consistent with the omnigenic framework's premise that peripheral genes, including non-coding elements, contribute to complex traits such as cancer.
Importantly, the inclusion of ncRNAs enhanced the identification of cancer-related pathways, indicating that a comprehensive network model encompassing ncRNAs is more effective in characterizing disease relationships than models concentrating solely on coding genes (\protect\hyperlink{ref-13VkGrWQR}{50}).

Empirical support for the omnigenic theory extends beyond human diseases.
Sundell et al.~demonstrated this concept in European black poplars, showing that both core and peripheral genes play a crucial role in predicting phenotypes.
While core genes are indeed important, the information they provide must be complemented by other genes to ensure accurate phenotype predictions.
Furthermore, peripheral genes were found to carry significant biological information contributing to robust predictions (\protect\hyperlink{ref-EgquVUSg}{51}).

@@

In essence, the omnigenic model proposes a specific mechanism: that the genetic architecture of complex traits is intrinsically interconnected, involving a multitude of genes that contribute to phenotypic variation both directly (core genes) and indirectly (peripheral genes).
This framework facilitates an understanding of the highly polygenic and often subtle genetic effects on complex traits, shifting the focus from individual genes to gene networks.
While this proves successful in some fields, the model exhibits certain limitations and has faced several criticisms.

@@

Furthermore, the omnigenic model has been criticized for its binary classification of genes into core and peripheral, which might oversimplify biological systems and potentially underestimate their true complexity (\protect\hyperlink{ref-82fs3SZo}{52}).
It also fails to account for gene-environment interactions, which play a crucial role in shaping complex traits (\protect\hyperlink{ref-dg9nKuy0}{55}).
Additionally, while the model provides a conceptual framework, it remains unclear how to translate it into a practical statistical model (\protect\hyperlink{ref-1GlQtJ9Yz}{56}).

@@

\begin{figure}
\hypertarget{fig:fig3}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/fig3.svg}
\caption{\textbf{Computational methods for machine learning derived gene module detection}.
\emph{(a)} Real biological network, where nodes represent genes and edges denote the relationships between them.
It captures all possible patterns that can link genes.
Some of these relationships are gene co-expression patterns, which transcriptomics can capture, such as (1), (3) and (4).
Other patterns, such as protein-protein interactions (2), are not captured by transcriptomics and are thus not reflected in the co-expression analysis.
\emph{(b)} Different approaches for capturing machine learning-derived gene modules.
In this example, transcriptome data is used as input, but other data types such as methylomics, proteomics, and metabolomics can also be applied.
The first row illustrates the Pathway-Level Information Extractor (PLIER) method, which uses non-negative matrix factorization and prior knowledge to align gene modules with known biological pathways.
This approach generates interpretable gene modules but only captures linear relationships.
The second row shows a variational autoencoders (VAEs), which can capture both linear and nonlinear patterns but does not integrate prior knowledge, resulting in less interpretable gene modules.
The third row highlights interpretable VAEs, which incorporate prior knowledge into its decoder, achieving a balance between capturing complex relationships and maintaining biological interpretability.}\label{fig:fig3}
}
\end{figure}

@@

GenomicSuperSignature employs PCA to identify gene modules applying prior knowledge for biological annotation after module discovery.
This flexibility allows it to adapt to novel datasets and provides computational efficiency even with large datasets.
However, the lack of prior knowledge integration during training can result in modules that are less biologically interpretable compared to those from PLIER-based methods (\protect\hyperlink{ref-X4fhSCkz}{65}).

@@

DL models are also highly effective in unsupervised settings such as inferring gene modules.
Specifically, DL methods like autoencoders (\protect\hyperlink{ref-NLVTJ9Lj}{73}), which utilize neural networks to learn compressed representations, can capture both linear and nonlinear patterns between genes.
This makes autoencoders particularly suited for uncovering gene modules and efficiently handling high-dimensional datasets.
For instance, variational autoencoders (VAEs) (\textbf{Figure \ref{fig:fig3}\emph{b}}) enhance this process by encoding data into a continuous latent space, enabling dimensionality reduction and the generation of new data.
However, this increased power to capture more general patterns has a major tradeoff: interpretability.
To overcome this, interpretable VAE models like the Pathway Module VAE (pmVAE) (\protect\hyperlink{ref-nGVsOd2S}{74}), expiMap (\protect\hyperlink{ref-JkqcQgM7}{75}), and VEGA (\protect\hyperlink{ref-170T6ip47}{12}) integrate biological pathway information into the decoder, which improves interpretability, thus making these approaches function as linear dimensionality reduction methods
(\textbf{Figure \ref{fig:fig3}\emph{b}}).

@@

\begin{figure}
\hypertarget{fig:fig4}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/fig4.svg}
\caption{\textbf{An integrative, gene module-based approach for genetic studies.}
Each panel shows a component of the PhenoPLIER framework (\protect\hyperlink{ref-NM3rHx1i}{13}).
\emph{(a)} First, latent variables (LVs) or gene modules are learned from transcriptome data using the Pathway-Level Information Extractor (PLIER) matrix factorization method.
PLIER generates matrix \(\mathbf{Z}\), which has gene weights for each module, and matrix \(\mathbf{B}\), which has the samples in the latent space.
\emph{(b)} Schematic of gene-trait associations from Transcriptome-wide association studies (TWAS) and gene-drug scores from LINCS L1000 being projected into the latent space for a joint analysis.
\emph{(c)} Schematic of a gene module-based drug reporposing framework, where the projection of TWAS and LINCS L1000 data is used to compute a drug-disease score.
\emph{(d)} Schematic of a regression model that tests whether genes that belong to a module (using a column of \(\mathbf{Z}\)) tend to be more strongly associated with a trait (using \(p\)-values from TWAS).
\emph{(e)} (top) Example of a gene module identified as LV246 analyzed in (\protect\hyperlink{ref-NM3rHx1i}{13}).
\emph{DGAT2} and \emph{ACACA}, found in a CRISPR screen to be linked to lipid metabolism and potential core genes, were among the top 15 genes in this module.
Using metadata from the top samples for LV246 from \(\mathbf{B}\), the module was found to be expressed mainly in adipose tissue, and liver and astrocyes were part of the top 10 cell types and tissues.
Using the regression model depicted in d), this module was significantly associated with high cholesterol, tryglycerides, LDL cholesterol, cholesterol lowering medication, Alzheimer's disease and dementia in the discovery cohort (UK Biobank), and high cholesterol and memory loss in the replication cohort (eMERGE).
(bottom) The \(p\)-values and colocalization probability from gene-trait associations from TWAS for a subset of the top 1\% genes in LV246 (\(y\)-axis) and traits significantly associated with LV246 (\(x\)-axis).
It can be seen that \emph{DGAT2} and \emph{ACACA} (in boldface), although strongly linked to lipid metabolism, are weakly associated with these lipid-relevant traits.}\label{fig:fig4}
}
\end{figure}

All the previously described approaches provide valuable and orthogonal information to better understand the molecular basis of human complex traits, including different molecular mechanisms through variant-trait, variant-gene, gene-trait and gene-gene information.
All of them have limitations and offer a partial view of the complex molecular interplay underlying pathophysiological processes.
In this section, we describe a gene module-based approach aimed at identifying patterns shared across these data sources.
The underlying hypothesis of this approach is that biologically meaningful, disease-relevant gene-gene links will be revealed when different yet incomplete data modalities are integrated.
In particular, we show examples on how the integration of gene modules with gene-trait associations and gene-drug data highlight core genes, disease-relevant molecular processes and drugs' mechanisms of action.
